Implementation and assessment of tools for integrating forest roads into projections of the cumulative effects of disturbance on wildlife

### 2024-05-17 analysis/paper/paper.Rmd
Sent to co-authors for review & comment.

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dev = "tiff",
  dev.args=list(compression="lzw"),
  dpi = 300
)
# Setting to avoid re-running projections
rerun <- FALSE
```

@@

To assess performance we calculated four characteristics of the projected and observed road networks that are relevant for wildlife: road density, measured as density of roads in meters per hectare (m/ha); distance to nearest road, measured as distance from each raster cell to the nearest road; road disturbance footprint, measured with a 1 km buffer around roads [@ibisch_global_2016]; and forestry disturbance footprint, measured with a 500 m buffer around roads and cutblocks. The forestry disturbance footprint is a variant of the buffered anthropogenic disturbance metric that is used as an indicator of boreal caribou critical habitat [@johnson_science_2020] and to quantify impacts of disturbance on Southern Mountain caribou [@lochhead_linking_2022; @polfus_identifying_2011]. The other metrics are also often used in studies of anthropogenic impacts on wildlife [e.g. @labadie_global_2023; @stewart_climate-informed_2023; @apps_factors_2006; @apps_spatial_2013; @serrouya_time_2017]. We also calculated road presence, measured as whether or not roads are present in each pixel, but we note that accurate projection of the exact locations of forest roads is not generally a requirement in stochastic ecological forecasting models wherein the exact locations of future disturbances are often not known. Forestry disturbance footprint and road density were calculated using the *disturbanceMetrics* and *rasterizeLineDensity* functions in the `caribouMetrics` R package [@hughes_cariboumetrics_2022]. Distance to nearest road was calculated using a fast parallel moving window method implemented in the *getDistFromSource* function of the `roads` package.

>>>>

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dev = "tiff",
  dev.args=list(compression="lzw"),
  dpi = 300
)
# Setting to avoid re-running projections
rerun <- TRUE
```

@@

To assess performance we calculated four characteristics of the projected and observed road networks that are relevant for wildlife: road density, measured as density of roads in meters per hectare (m/ha); distance to nearest road, measured as distance from each raster cell to the nearest road; road disturbance footprint, measured with a 1 km buffer around roads [@ibisch_global_2016]; and forestry disturbance footprint, measured with a 500 m buffer around roads and cutblocks. The forestry disturbance footprint is a variant of the buffered anthropogenic disturbance metric that is used as an indicator of boreal caribou critical habitat [@johnson_science_2020] and to quantify impacts of disturbance on Southern Mountain caribou [@lochhead_linking_2022; @polfus_identifying_2011]. The other metrics are also often used in studies of anthropogenic impacts on wildlife [e.g. @labadie_global_2023; @stewart_climate-informed_2023; @apps_factors_2006; @apps_spatial_2013; @serrouya_time_2017]. We also calculated road presence, measured as whether or not roads are present in each pixel, but we note that accurate projection of the exact locations of forest roads is not generally a requirement in stochastic ecological forecasting models wherein the exact locations of future disturbances are often not known. Forestry disturbance footprint and road density were calculated using the *disturbanceMetrics* and *rasterizeLineDensity* functions in the `caribouMetrics` R package [@hughes_cariboumetrics_2022].

### 2024-05-17 analysis/paper/supplement.Rmd
Sent to co-authors for review & comment.

Profiling with the `Rprof` function shows which parts of a function take the longest. On a test run with 1 million cells in the weight raster and 10 landings we profiled the performance of the ILCP and MST methods. Both methods begin by constructing a full graph, which took 3-5 s. For the ILCP method, iteratively determining the shortest paths is the only other step that takes more than a second to run, and this run time will increase with the number of landings (Figure \@ref(fig:profiling-ilcp)). For the MST method, calculating cost distances among landings and from each landing to the nearest point on the existing road distance takes the most time, which also increases with the number of landings (Figure \@ref(fig:profiling-mst)). Determining the locations of the paths in the minimum spanning tree also takes about 6 seconds. We also considered using the `gDistance` package [@van_etten_r_2017] to build the graph but it took longer than our method (Figure \@ref(fig:profiling-ilcp-gdistance)).

```{r profiling-ilcp, fig.cap="Profiling results for the ILCP method showing functions that take > 1 s to run on a cost raster with 1 million cells and 10 landings. Functions higher in the call stack are called by functions lower down.", fig.height=2}
if(!file.exists(here::here("analysis/data/derived_data/ilcp_100_Rprof.out"))){
  source(here::here("analysis/scripts/test_profile.R"), keep.source = TRUE)
}

>>>>

Profiling with the `Rprof` function shows which parts of a function take the longest. On a test run with 1 million cells in the weight raster and 10 landings we profiled the performance of the ILCP and MST methods. Both methods begin by constructing a full graph, which took 3-5 s. For the ILCP method, iteratively determining the shortest paths is the only other step that takes more than a second to run, and this run time will increase with the number of landings (Figure \@ref(fig:profiling-ilcp)). For the MST method, calculating cost distances among landings and from each landing to the nearest point on the existing road distance takes the most time, which also increases with the number of landings (Figure \@ref(fig:profiling-mst)). Determining the locations of the paths in the minimum spanning tree also takes about 6 seconds. We experimented with using graph construction methods in the package `gdistance` [@van_etten_r_2017], but Lochhead et al's [-@lochhead_demo_2018; -@lochhead_linking_2022] method is faster (compare Figures \@ref(fig:profiling-ilcp) and \@ref(fig:profiling-ilcp-gdistance)).

```{r profiling-ilcp, fig.cap="Profiling results for the ILCP method showing functions that take > 1 s to run on a cost raster with 1 million cells and 10 landings. Functions higher in the call stack are called by functions lower down.", fig.height=3}
if(!file.exists(here::here("analysis/data/derived_data/ilcp_100_Rprof.out"))){
  source(here::here("analysis/scripts/test_profile.R"), keep.source = TRUE)
}
